---
title: "week2-prj"
author: "Ravi"
date: "`r Sys.Date()`"
output: html_document
---

------------------------------------------------------------------------

Reproducible Research: Peer Assessment 1 \## Loading and pre-processing the data

First, we load the data and convert the `date` column to `Date` format.

```{r load-data, echo=TRUE}
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
```

We calculate the total number of steps taken per day and then create a histogram of these totals. We also calculate and report the mean and median total number of steps taken per day.

```{r step1}
# Total steps per day
total_steps_per_day <- tapply(data$steps, data$date, sum, na.rm=TRUE)

# Histogram of total steps per day
hist(total_steps_per_day, main="Total Steps per Day", xlab="Steps", ylab="Frequency", col="blue")

# Mean and median of total steps per day
mean_steps <- mean(total_steps_per_day, na.rm=TRUE)
median_steps <- median(total_steps_per_day, na.rm=TRUE)
mean_steps
median_steps
```

We calculate the average number of steps taken in each 5-minute interval, across all days, and create a time series plot.

```{r step2}
# Average steps per 5-minute interval
avg_steps_per_interval <- tapply(data$steps, data$interval, mean, na.rm=TRUE)

# Time series plot of average steps per interval
plot(names(avg_steps_per_interval), avg_steps_per_interval, type="l", 
     main="Average Steps per Interval", xlab="Interval", ylab="Average Steps")

# 5-minute interval with maximum average steps
max_interval <- names(avg_steps_per_interval)[which.max(avg_steps_per_interval)]
max_interval
```

Imputing missing values

```{r step3}
# Total number of missing values
missing_values <- sum(is.na(data$steps))
missing_values

# Impute missing values
imputed_data <- data
for (i in 1:nrow(imputed_data)) {
  if (is.na(imputed_data$steps[i])) {
    interval <- imputed_data$interval[i]
    imputed_data$steps[i] <- avg_steps_per_interval[as.character(interval)]
  }
}
# Total steps per day after imputation
total_steps_per_day_imputed <- tapply(imputed_data$steps, imputed_data$date, sum)

# Histogram of total steps per day (imputed data)
hist(total_steps_per_day_imputed, main="Total Steps per Day (Imputed)", xlab="Steps", ylab="Frequency", col="green")

# Mean and median of total steps per day (imputed data)
mean_steps_imputed <- mean(total_steps_per_day_imputed)
median_steps_imputed <- median(total_steps_per_day_imputed)
mean_steps_imputed
median_steps_imputed
```

We create a new factor variable in the dataset with two levels -- "weekday" and "weekend".

```{r step4}
# Create a new factor variable for weekday/weekend
imputed_data$weekday <- ifelse(weekdays(imputed_data$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
```

We create a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.

```{r step5}
# Average steps per interval for weekdays and weekends
avg_steps_weekday <- tapply(imputed_data$steps[imputed_data$weekday == "weekday"], 
                            imputed_data$interval[imputed_data$weekday == "weekday"], mean)
avg_steps_weekend <- tapply(imputed_data$steps[imputed_data$weekday == "weekend"], 
                            imputed_data$interval[imputed_data$weekday == "weekend"], mean)

# Panel plot comparing weekdays and weekends
par(mfrow=c(2,1))
plot(names(avg_steps_weekday), avg_steps_weekday, type="l", 
     main="Average Steps per Interval (Weekday)", xlab="Interval", ylab="Average Steps")
plot(names(avg_steps_weekend), avg_steps_weekend, type="l", 
     main="Average Steps per Interval (Weekend)", xlab="Interval", ylab="Average Steps")
```
